Legacy Data Confounds Genomics Studies

### 2019-02-19 legacy.tex
cleaned up misc hashed out text
small edits

\sgcomment{Check time concordance throughout}		
%\subsection{Batch Effects in Aging Reference Cohort Data}

@@

Even though such batch effects clearly have the potential to confound analyses, the usefulness and ubiquity of the 1kGP data make such issues particularly insidious.   
This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
Here we point out how large and previously unreported batch effects in the early phases of the 1kGP still lead to incorrect genetic conclusions through population genetic analyses \luke{and spurious GWAS associations as a result of imputation using the 1kGP as a reference.}

@@

Some mutagenic processes will affect different DNA motifs. 
Certain mutagens in tobacco smoke, for example, have been shown to preferentially bind to certain genomic motifs leading to an excess of G to T transversions \citep{Pfeifer2002,Pleasance2010}. 
Thus, exposure of populations to different mutational processes can be inferred by considering the DNA context of polymorphism in search of \textit{signatures} of different mutational processes \citep{Alexandrov2013,Shiraishi2015a}. 
Such genome-wide mutational signatures have been used as diagnostic tools for cancers \citep{Alexandrov2013,Shiraishi2015a}.

@@

Harris and Pritchard reported an excess of a 3-mer substitution patterns *AC${\rightarrow}$*CC in a portion of the Japanese individuals in the 1kGP \citep{Harris2017a}.
While trying to follow up on this observation in a larger and more recent Japanese cohort, we did not find this particular signature.
When comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed a number of single nucleotide polymorphisms (SNPs) private to one of the two groups.
Given the similarity of the two populations, this strongly suggests a technical difference rather than a population structure effect.
These mismatches were maintained after filtering for low-quality regions of the human genome and sites failing Hardy-Weinberg equilibrium tests. 
%\sgcomment{did we have a specific result for this? As in "$94\%$ of sites passed HW"...}. 
%Even though some of the mismatched variants are at low frequency in the 1kGP, and thus can evade Hardy-Weinberg detection, but many are at high enough frequency that HW detection   
%Whereas Hardy-Weinberg testing is efficient at detecting high frequency spurious variants, it has little power to filter out rare variants which, in aggregate, can contribute to mutational signatures.

When mismatch sites are removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears (Figure \ref{SFS}). To identify possible technical reasons for the difference, we performed regressions of the prevalence of the  *AC${\rightarrow}$*CC mutational signature against different individual-level quality metrics provided by the 1kGP (see Supplementary Figure \ref{PC1_Correlation}). The average quality of mapped bases  $Q$ per individual stood out as a strong correlate : Individuals with low $Q$ show elevated rates of the signature. %(\sgcomment {p value?})
Thus, sequences with low mapping quality harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.

@@

Thus, this mutational signal is heavily enriched in suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques (Figure \ref{MutSpect}).
The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged (Figure \ref{MutSpect}).
For population genetic analyses sensitive to the accumulation of rare variants, the removal of individuals with low $Q$ appears preferable to filtering specific low-quality SNPs. 
For other analyses, removing suspicious SNPs may be adequate. 
Lists of suspicious SNPs and low-quality individuals is provided as supplementary material \luke{link to file here}.

\subsection{Identifying suspicious variants in the 1kGP}
The distribution of $Q$ across 1kGP populations shows that many populations have distributions of $Q$ scores comparable to that of the JPT, especially populations sequenced in the phase 1 of the project: sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
This variability could result from evolving sequence platform and protocols or variation between sequencing centres. 
By 2011, older sequencing technologies were phased out, and methods became more consistent, resulting in higher and more uniform quality.
%Among the many covariates that could be used to identify spurious sites, such as sequencing date, sequencing technology, or sequencing centre, we used average quality of mapped bases ($Q$) per individual because it best correlated with mismatched sites across Japanese cohorts.

We therefore performed the same reverse GWAS approach in all populations independently, and similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
Over 3826 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ (Figure \ref{OverLap}).

@@

We find that of the 1899 variants from the full model that are present in both Resequenced and original 1kGP data, 933 of the variants are present in both samples at matching allele frequencies.
These variants are likely unaffected by $Q$ in these particular individuals.
We find that 772 variants are at higher frequency in the resequenced data.
These variants may be associated to $Q$ due to false negatives, whereby individuals with lower $Q$ tend to have lower calling rates for these variants.
Only 194 variants are at higher frequencies in the original 1kGP data.
These variants may be associated to $Q$ due to false positives, whereby individuals with lower $Q$ have higher calling rates for these variants (See Figure \ref{90HanSFS_full}).

@@

%\begin{figure}[h]
%\includegraphics[width=\hsize,keepaspectratio]{./Figures/AllDeviances.jpg}

%\caption{Empirical distribution plots of the sum of deviances for Non-Repeat variants grouped by N, the number of populations included. The blue curve is the theoretical null reference chi squared distribution with $N$ degrees of freedom. Note $\sum_i^N  \chi^2_1= \chi^2_N$. The grey dotted line indicates the significance level of $10^{-5}$ for the respective distribution which happens to be very close to the adjusted \textit{p}-value threshold at 0.01 FDR. }
%\label{Deviances}
%\end{figure}

@@

%\begin{figure}[h]
%\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot_mostSig2.jpg}
%\caption{Logistic regression of the most significantly associated SNP rs75254682 for the populations with significant association to the average quality per mapped bases $Q$.}
%\label{MostSig}
%\end{figure}

%\begin{figure}[h]
%\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot_diabetes.jpg}
%\caption{Logistic regression of rs301 to the average quality per mapped bases $Q$.}
%\label{DiabetesSNP}
%\end{figure}

%\begin{figure}[h]
%\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot.jpg}
%\caption{Logistic regression of rs6057648 for the populations with significant association to the average quality per mapped bases $Q$.}
%\label{TwinsSNP}
%\end{figure}

%\begin{figure}[h]
%\includegraphics[width=\hsize,keepaspectratio]{./Figures/SNP6_Singles.jpg}
%\caption{SNPs associated with $Q$ in a single population. The x-axis is sorted by genomic position.}
%\label{Singles}
%\end{figure}

>>>>

\subsection{Batch Effects in Aging Reference Cohort Data}

@@

Even though such batch effects clearly have the potential to confound analyses, the unrestricted availability of the 1kGP data lead to ubiquitous use making batch effects particularly insidious.   
This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
Here we point out how large and previously unreported batch effects in the early phases of the 1kGP still lead to incorrect genetic conclusions through population genetic analyses \luke{and spurious GWAS associations as a result of imputation using the 1kGP as a reference.}

@@

Different mutagenic processes may preferentially affect different DNA motifs. 
Certain mutagens in tobacco smoke, for example, have been shown to preferentially bind to certain genomic motifs leading to an excess of G to T transversions \citep{Pfeifer2002,Pleasance2010}. 
Thus, exposure of populations to different mutational processes can be inferred by considering the DNA context of polymorphism in search of \textit{signatures} of different mutational processes \citep{Alexandrov2013,Shiraishi2015a}. 
Such genome-wide mutational signatures have been used as diagnostic tools for cancers \citep{Alexandrov2013,Shiraishi2015a}.

@@

Harris and Pritchard reported an excess of a 3-mer substitution patterns *AC${\rightarrow}$*CC in a portion of the Japanese individuals in the 1kGP \citep{Harris2017a}.
While trying to follow up on this observation in a larger and more recent Japanese cohort, we did not find this particular signature.
When comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed a number of single nucleotide polymorphisms (SNPs) private to one of the two groups.
Given the similarity of the two populations, this strongly suggests a technical difference rather than a population structure effect.
These mismatches were present despite filtering for low-quality regions of the human genome and Hardy-Weinberg disequilibrium.

When mismatch sites are removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears (Figure \ref{SFS}). To identify possible technical reasons for the difference, we performed regressions of the prevalence of the  *AC${\rightarrow}$*CC mutational signature against different individual-level quality metrics provided by the 1kGP (see Supplementary Figure \ref{PC1_Correlation}). The average quality of mapped bases  $Q$ per individual stood out as a strong correlate : Individuals with low $Q$ show elevated rates of the signature. %(\sgcomment {p value?})
Thus, sequences called from low-$Q$ data contain mutations that reproduce poorly across studies and exhibit a particular mutational signature.

@@

Thus, this mutational signal is heavily enriched in suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques (Figure \ref{MutSpect}).
The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged (Figure \ref{MutSpect}).
For population genetic analyses sensitive to the accumulation of rare variants, the removal of individuals with low $Q$ appears preferable to filtering specific low-quality SNPs. 
For other analyses where quality of imputation matters, identifying suspicious SNPs may be preferable. 
Lists of suspicious SNPs and individuals with low $Q$ are provided as supplementary material \luke{link to file here}.

\subsection{Identifying suspicious variants in the 1kGP}
The distribution of $Q$ across 1kGP populations shows that many populations have distributions of $Q$ scores comparable to that of the JPT, especially populations sequenced in the phase 1 of the project: sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
This variability could result from evolving sequence platform and protocols or variation between sequencing centres. 
By 2011, older sequencing technologies were phased out, and methods became more consistent, resulting in higher and more uniform quality.

We therefore performed the same reverse GWAS approach in all populations independently, and similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
Over 3826 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ in each (Figure \ref{OverLap}).

@@

Of the 1899 variants from the full model that are present in both Resequenced and original 1kGP data, 933 of the variants are present in both samples at matching allele frequencies.
These variants are likely unaffected by $Q$ in these particular individuals.
We find that 772 variants are at higher frequency in the resequenced data.
These variants may be associated to $Q$ due to false negatives, whereby individuals with lower $Q$ tend to have lower calling rates for these variants.
Only 194 variants are at higher frequencies in the original 1kGP data.
These variants may be associated to $Q$ due to false positives, whereby individuals with lower $Q$ have higher calling rates for these variants (See Figure \ref{90HanSFS_full}).

@@

@@
